Manuscript_draft

### 2021-01-03 manuscript_draft.Rmd
manny commit

### Stand-level volume and basal area estimates
Total and merchantable volumes were estimated for trees using species- and region-specific equations [@nigh2016] based on tree height and DBH.  Consistent with forest management practices in British Columbia, we estimated volumes for trees meeting a minimum diameter utilization limit of 17.5cm DBH.  For trees lacking a tree height measurement, we estimated height by fitting a single diameter-height model developed with a linear mixed effects model, with log-transformed DBH as a fixed effect, and tree nested within plot as a random effect.  Trees with broken tops were omitted from diameter-height models. We summed tree-level estimates of volume and basal area to generate stand-level estimates for each plot. To test the hypothesis that changes in stand attributes over 27 years varied by treatment, we fit separate linear models with stand total volume, basal area, and density as response variables, and the interaction between treatment and year as the independent variable.

@@

  # now summarize at treatment unit level
  group_by(BA.Target,Period) %>% 
   summarise(Vol=mean(Vol.sum,na.rm=T),
            Vol.se=incSE(Vol.sum),
            BA=mean(BA.sum,na.rm=T),
            BA.se=incSE(BA.sum),
            SPH=mean(SPH.sum,na.rm=T),
            SPH.se=incSE(SPH.sum),
            QMD=mean(QMD.mean,na.rm=T),
            QMD.se=incSE(QMD.mean))

@@

All units increased volume, basal area and stand density over the 1992-2019 period (`r figNums("standPlots",display="cite")`).  The rate of change differed between units.  The low RBA unit accrued basal area faster than the other two units over the 1992-2019 period, and this difference was significant (`r appNums("standFit.contrasts",display="cite")`).  The estimated rate of basal area increment in the low RBA unit was higher by 
`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Basal area") %>%   pull(estimate) %>% round(.,1) %>%  abs() %>%   .[1]`
and
`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Basal area") %>%   pull(estimate) %>% round(.,1) %>%  abs() %>%   .[2]`
m² per hectare more than the high RBA and control units, respectively (p<0.01 for both). The low RBA unit also significantly increased its stand density by an estimated 
`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Stand density") %>%   pull(estimate) %>% round(.,1) %>%  abs() %>%   .[2]`
stems per hectare more than the control unit (p<0.001). The rate of change in tree density did not differ significantly between the two harvested units.

The rate of change in volume in trees over 18.5cm DBH was different between barely significant between the two harvested units, and there were no significant differences between the control and harvested units.

@@

@@

@@

periodInc.plot<-function(standVar,y.label){

  perInc.X %>%
  dplyr::select(BA.Target,Period,standVar,paste0(standVar,".se")) %>% 
    setNames(c("BA.Target","Period","Mean","SE")) %>%

  ggplot(aes(x=Period,y=Mean,group=BA.Target,color=BA.Target))+
  geom_point()+
  geom_line()+
     geom_errorbar(aes(ymin=Mean-SE, ymax=Mean+SE, width=.2),
                   position=position_dodge(0)) %>% 
    return()
}

periodInc.plot("BA",y.label="BA (m²/ha)")
periodInc.plot("Vol",y.label="BA (m²/ha)")

@@

>>>>

### Stand-level volume and basal area estimates
Total and merchantable volumes were estimated for trees using species- and region-specific equations [@nigh2016] based on tree height and DBH.  Consistent with forest management practices in British Columbia, we estimated volumes for trees meeting a minimum diameter utilization limit of 18.5cm DBH.  For trees lacking a tree height measurement, we estimated height by fitting a single diameter-height model developed with a linear mixed effects model, with log-transformed DBH as a fixed effect, and tree nested within plot as a random effect.  Trees with broken tops were omitted from diameter-height models. We summed tree-level estimates of volume and basal area to generate stand-level estimates for each plot. To test the hypothesis that changes in stand attributes over 27 years varied by treatment, we fit separate linear models with stand total volume, basal area, and density as response variables, and the interaction between treatment and year as the independent variable.

@@

  # now summarize at treatment unit level
  group_by(BA.Target,Period) %>% 
   summarise(Vol=mean(Vol.sum,na.rm=T),
            Vol.se=incSE(Vol.sum),
            BA=mean(BA.sum,na.rm=T),
            BA.se=incSE(BA.sum),
            SPH=mean(SPH.sum,na.rm=T),
            SPH.se=incSE(SPH.sum),
            QMD=mean(QMD.mean,na.rm=T),
            QMD.se=incSE(QMD.mean)) %>%

  # rename periods
  mutate(Period=dplyr::recode(Period,
                       Period1="1992-1994",
                       Period2="1994-1997",
                       Period3="1997-2009",
                       Period4="2009-2019"))

@@

All units increased volume, basal area and stand density over the 1992-2019 period (`r figNums("standPlots",display="cite")`).  Mean annual basal area and volume increment peaked during the 1997-2009 period in the two harvested treatments, and peaked in the 1994-1997 period in the control treatment unit (`r figNums("perInc.plot",display="cite")`.  The low RBA unit accrued basal area faster than the other two units over the 1992-2019 period, and this difference was significant (`r appNums("standFit.contrasts",display="cite")`).  The estimated rate of basal area increment in the low RBA unit was higher by 
`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Basal area") %>%   pull(estimate) %>% round(.,1) %>%  abs() %>%   .[1]`
and
`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Basal area") %>%   pull(estimate) %>% round(.,1) %>%  abs() %>%   .[2]`
m² per hectare more than the high RBA and control units, respectively (p<0.01 for both). The low RBA unit also significantly increased its stand density by an estimated 
`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Stand density") %>%   pull(estimate) %>% round(.,1) %>%  abs() %>%   .[2]`
stems per hectare more than the control unit (p<0.001). The rate of change in tree density did not differ significantly between the two harvested units.  The rate of change in volume in trees over 18.5cm DBH was different between barely significant between the two harvested units, and there were no significant differences between the control and harvested units.

@@

## Mortality

@@

## Management considerations
1. Both species responded with increased growth after partial harvest.  High removal of basal area in these forests stimulated stronger spruce growth.  This information can help silviculturists manipulate stand structure to preferentially ...
2. Partial harvesting at both intensities delayed peak volume and basal area increment by approximately 10 years.  Annual increment in the control unit declined faster over the last measurement period compared to previous periods, perhaps suggesting that the stand is becoming mature and losses due to mortality may substantially offset growth increment.

@@

# Note this function performs the operation for basal area and volume only

  perInc.X %>%
  dplyr::select(BA.Target:BA.se) %>% 
    setNames(c("BA.Target","Period","Vol_mean","Vol_SE","BA_mean","BA_SE")) %>% 
    pivot_longer(cols=-c(BA.Target,Period),
                 names_to=c("Var","SE"),
                 names_sep="_",
                 values_to="Value") %>% 
    pivot_wider(names_from="SE",values_from="Value") %>%

  ggplot(aes(x=Period,y=mean,group=BA.Target,color=BA.Target))+
  geom_point()+
  geom_line()+

     geom_errorbar(aes(ymin=mean-SE, ymax=mean+SE, width=.2),
                   position=position_dodge(0))+
    facet_wrap(~Var,nrow=2,
               scales="free_y", strip.position = "left",
               labeller=as_labeller(c(BA="Basal area inc. (m²/ha/year)",Vol="Volume inc. (m²/ha/year)")))+
    ylab(NULL)+
     theme(strip.background = element_blank(),
           strip.placement = "outside")

@@

`r figNums("perInc.plot",caption="Mean plot annual increment over four measurement periods for basal area and volume, by treatment.  Whiskers are one standard error of the mean. Inc, increment.")`
